---
title: "ESM 244 Lab Lab 2"
author: "Katelin Seeto"
date: "1/14/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set

library(tidyverse)
library(janitor)
library(palmerpenguins)
library(here)

# For PCA
library(ggfortify)

# For ggplot customization
library(readxl)
library(gghighlight)
library(lubridate)
library(plotly)

# For compound figures 
library(patchwork)
```

## Part 1: PCA with penguins data

First some wrangling! Select columns we need, remove n/a since isn't useful when looking at multivariate space, scale data to make sure no variable is over weighted in principle components just due to the units it is measured in. 

Notice that body mass is in grams and values are in 1000's compared with bill length in mm with values in 10's

Notes:  
* `ends_with()` is a helper function to select all variables ending with a certain string  
* `drop_na()` drops all rows with n/a values, can write variables inside () to specifify which columns to drop n/a from  
* `scale()` scales the data  
* `prcomp()` runs principle components and changes your df into a list 
* `autoplot()` uses ggplot2 to draw a particular plot for an object of a particular class in a single command. For ex, for PCA data type it will assume you want a PCA biplot

```{r}
penguins_pca <- penguins %>% 
  select(body_mass_g, ends_with("_mm")) %>% 
  drop_na() %>% 
  scale() %>% 
  prcomp()

penguins_pca$rotation # brings up the loadings for each of the 4 variables along that principle component

# Create biplot with autoplot(), doesn't have info about variable loadings, or label penguin spp!
autoplot(penguins_pca)

# Create new plot which contains info we can update aesthetics by, like spp and other variables
penguin_complete <- penguins %>% 
  drop_na(body_mass_g, ends_with("_mm"))
# Includes PCA data, and the observations used to make that PCA 
# Observations used to create PCA and data used for aesthetics MUST align!
autoplot(penguins_pca,
         data = penguin_complete, 
         colour = 'species',
         loadings = TRUE, 
         loadings.label = TRUE)+
  theme_minimal()

```

## Part 2: ggplot2 cutomization and reading in different file types 

### Read in an .xlsx file and do some wrangling 

Notes:  
* `clean_names()` default is to change all column headings to lower snake case
* `mutate()` can also be used to transform existing columns, not just to create new ones  
* `across()` is a helper function, can be used to say like across last 3 cols, or cols that end with...  
* `tolower()` transform to lowercase 
* `str_sub()` extracts or replaces substrings (part of a string) from a character vector

```{r}
fish_noaa <- read_excel(here("data", "foss_landings.xlsx")) %>% 
  clean_names() %>% 
  mutate(across(where(is.character), tolower)) %>% # Across any column where that col is a character, want to use function tolower() to change to all lowercase  
  mutate(nmfs_name = str_sub(nmfs_name, end = -4)) %>% # overwrites col since name nmfs_name is same name as existing column
  filter(confidentiality == "public")
```

### Make a customized graph: 

Notes:  
* `ggplotly()` create interactive graph  
* Can highlight using `gghighlight()` to highlight certain series or values
```{r}
fish_plot <- ggplot(data = fish_noaa, aes(x = year, y = pounds))+ 
  geom_line(aes(color = nmfs_name), show.legend = FALSE)+ 
  theme_minimal() 

fish_plot

ggplotly(fish_plot)

# Use gghighlight to highlight certain series 
ggplot(data = fish_noaa, aes(x = year, y = pounds, group = nmfs_name))+ 
  geom_line()+ 
  theme_minimal()+ 
  gghighlight(nmfs_name == "tunas")

# Use gghighlights to highlight certain values 
ggplot(data = fish_noaa, aes(x = year, y = pounds, group = nmfs_name))+ 
  geom_line(aes(color = nmfs_name))+ 
  theme_minimal()+ 
  gghighlight(max(pounds) > 1e8)


```

### Read in data from a URL, use `lubridate()`, `mutate()`, make a graph with months in logical order

Notes:  
* Can use `read_csv()` using url links, but must weigh benefits/ costs of not knowing exactly what the dataset looked like when you did analysis. May want to download hard copy  
* `mdy()` from lubridate to convert to date  
* `month.abb[]` function in base R to replace abbreviation of month name by number, `month.name[]` replaces with full name, could use `case_when()` to manually do the same 

```{r}
monroe_wt <- read_csv("https://data.bloomington.in.gov/dataset/2c81cfe3-62c2-46ed-8fcf-83c1880301d1/resource/13c8f7aa-af51-4008-80a9-56415c7c931e/download/mwtpdailyelectricitybclear.csv")
```

```{r}
monroe_ts <- monroe_wt %>% 
  mutate(date = mdy(date)) %>% 
  mutate(record_month = month(date)) %>% #creates new column with just the month from the date column
  mutate(month_name = month.abb[record_month]) %>% # Create month name and not just the month number 
  mutate(month_name = fct_reorder(month_name, record_month))# change month to be an ordered factor so it won't show up alphabetically in ggplot 

ggplot(data = monroe_ts, aes(x = month_name, y = total_kWh))+
  geom_jitter()
```









